Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-03-23 paper/paper.tex
Change words
More writing

\newcommand{\lineref}[1]{line~\ref{#1}}

@@

\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\textfraction}{0.05}
\renewcommand{\floatpagefraction}{0.85}

@@

@@

?? Stuff on EHPs, how work stealing matters here too

@@

@@

\begin{algorithm}[tb] \DontPrintSemicolon
\nl $\FuncSty{associationMCCIS}$ :: (Graph $G_1$, Graph $G_2$) $\rightarrow$ Map \;
\nl \Begin{
    \nl $\KwSty{global} ~ \mathit{incumbent} \gets \emptyset$ \;
    \nl $G \gets \FuncSty{encode}(G_1, G_2)$ \;
    \nl $\FuncSty{search}(G, \emptyset, \emptyset, \operatorname{V}(G))$ \;
    \nl $\KwSty{return} ~ \FuncSty{decode}(C_{\mathit{max}})$ \;
}
\vspace{1ex}
\nl $\FuncSty{search}$ :: (Graph $G$, Set $\mathit{solution}$, Set $\mathit{branchable}$, Set $\mathit{remaining}$) \;
\nl \Begin{
    \nl $\mathit{colourClasses} \gets \FuncSty{concatenate}($ \;
    $\hspace*{2em}\FuncSty{colour}(G, \mathit{remaining} \setminus \mathit{branchable}),$ \;
    $\hspace*{2em}\FuncSty{colour}(G, \mathit{remaining} \cap \mathit{branchable}))$ \;
    \nl \While{$\FuncSty{length}(\mathit{colourClasses}) > 0$}{
        \nl \ForEach{$v \in \FuncSty{last}(\mathit{colourClasses}) \textnormal{~in reverse order}$}{
            \nl \lIf{$|\mathit{solution}| + \FuncSty{length}(\mathit{colourClasses}) \leq |\mathit{incumbent}|$}{\KwSty{return}}
            \nl \lIf{$v \notin \mathit{branchable} ~ \KwSty{and} ~ \mathit{solution} \ne \emptyset$}{\KwSty{return}}
            \nl $\mathit{solution}' \gets \mathit{solution} \cup \{ v \}$\;
            \nl \lIf{$|\mathit{solution}'| > |\mathit{incumbent}|$}{$\mathit{incumbent} \gets
            \mathit{solution}'$}
            \nl $\mathit{branchable}' \gets \mathit{branchable} \cup \FuncSty{connected}(v)$ \;
            \nl $\mathit{remaining}' \gets \mathit{remaining} \cap \operatorname{N}(G, v)$ \;
            \nl \lIf{$\mathit{remaining}' \neq \emptyset$}{$\FuncSty{search}(G, \mathit{solution}',
            \mathit{branchable}', \mathit{remaining}')$}
         }
         \nl $\FuncSty{removeLast}(\mathit{colourClasses})$ \;
     }
}
\vspace{1ex}
\nl $\FuncSty{colour}$ :: (Graph $G$, Set $\mathit{uncoloured}$) $\rightarrow$ List of List of Vertex \;
\nl \Begin{
    \nl $\mathit{result} \gets []$ \;
    \nl \While{$\mathit{uncoloured} \ne \emptyset$}{
        \nl $\mathit{colourClass} \gets []$ \;
        \nl $\mathit{colourable} \gets \mathit{uncoloured}$ \;
        \nl \While{$\mathit{colourable} \ne \emptyset$}{
            \nl $v \gets \textnormal{the first vertex from}~\mathit{uncoloured}~\textnormal{by
            static degree order from}~G$ \;
            \nl $\FuncSty{append}(\mathit{colourClass}, v)$ \;
            \nl $uncoloured \gets uncoloured \setminus \{ v \}$ \;
            \nl $\mathit{colourable} \gets \mathit{colourable} \setminus \operatorname{N}(G, v)$ \;
        }
        \nl $\FuncSty{append}(\mathit{result}, \mathit{colourClass})$ \;
    }
    \nl $\KwSty{return} ~ \mathit{result}$ \;
}
\caption{An algorithm for MCCIS via an association graph.}
\label{algorithm:mc}
\end{algorithm}

We can use the branching idea inside a ``maximum clique'' algorithm, by providing additional
information on the meaning behind vertices in the association graph as input. To fit in with the
bitset encoding, we preconstruct a bitset for each vertex $v$ in the association graph, saying which
association graph vertices correspond to vertices adjacent to $v$ in the left-hand graph.

?? This gets confusing. Maybe we can come up with some clever terminology: when I was talking about
search trees and clique algorithms, I used the term 'node' for something in a search tree, and
'vertex' for something in a graph. Maybe we can use different terms for things in the association
graph, versus things in the two input graphs? Should we call association graph vertices
'assignments' and edges 'compatibilities' instead?

We keep track of allowed branching vertices in a set we call $A$. We only branch on vertices which
are present in $P \cap A$.

?? We could make this faster, since it'll be the same for each vertex corresponding to the same
left-hand vertex.

We must also ensure that the bound remains valid. The branching and bounding are integrated, and it
is not possible to alter the branching rule without also changing the bound. Our observation is that
the bound rule works for any valid colouring, not just a greedy one, so long as we always iterate
from the last colour class downwards, without skipping any vertices. Thus we construct a two-pass
greedy colouring such that all the vertices in $P \setminus A$ are to the left, and the vertices in
$P \cap A$ are to the right.

\cite{DBLP:conf/cp/McCreeshP14}

>>>>

\newcommand{\lineref}[1]{line~\ref{#1}}
\newcommand{\twolinesref}[2]{lines~\ref{#1} and~\ref{#2}}

@@

@@

A \emph{graph} $G = (V, E)$ consists of a set of vertices $V$ and a set of edges $E \subseteq V
\times V$. In this paper our graphs may be directed, may have loops, and there is a label associated
with each vertex and with each edge. For convenience, we treat undirected graphs as being directed
graphs with symmetric adjacencies, and for unlabelled graphs we treat each vertex as having the same
label, and each edge as having the same label. We write $\operatorname{V}(G)$ for $V$, and
$\operatorname{N}(G, v)$ for the \emph{undirected neighbourhood} of a vertex $v$ (that is, the set
of vertices $w$ such that either $(v, w)$ or $(w, v)$ is an edge).

Given two graphs $G_1$ and $G_2$, we say two pairs of vertices $(v_1, w_1) \in \operatorname{V}(G_1)
\times \operatorname{V}(G_1)$ and $(v_2, w_2) \in \operatorname{V}(G_2) \times
\operatorname{V}(G_2)$ are \emph{compatible} if both $v_1$ and $w_1$ and $v_2$ and $w_2$ are either
non-adjacent or adjacent with the same label, and if the same holds for $(w_1, v_1)$ and $(w_2,
v_2)$.  We say two vertices $v \in \operatorname{V}(G_1)$ and $w \in \operatorname{V}(G_2)$ are
\emph{compatible} if they share the same label, and $(v, v)$ is compatible with $(w, w)$.

@@

?? Stuff on EHPs, how work stealing matters here too \cite{DBLP:journals/jco/BatsynGMP14}

@@

?? We need a picture here.

@@

\begin{algorithm}[p] \DontPrintSemicolon
\nl $\FuncSty{associationMCCIS}$ :: (Graph $G_1$, Graph $G_2$) $\rightarrow$ Map \;
\nl \Begin{
    \nl $\KwSty{global} ~ \mathit{incumbent} \gets \emptyset$ \label{incumbent} \;
    \nl $G \gets \FuncSty{associationGraph}(G_1, G_2)$ \label{buildassoc} \;
    \nl $\FuncSty{search}(G, \emptyset, \emptyset, \operatorname{V}(G))$ \label{firstsearch} \;
    \nl $\KwSty{return} ~ \mathit{incumbent}$ \;
}
\vspace{1ex}
\nl $\FuncSty{search}$ :: (Graph $G$, Set $\mathit{solution}$, Set $\mathit{branchable}$, Set $\mathit{remaining}$) \;
\nl \Begin{
    \nl $\mathit{colourClasses} \gets \FuncSty{concatenate}($ \label{makecolours} \;
    $\hspace*{2em}\FuncSty{colour}(G, \mathit{remaining} \setminus \mathit{branchable}),$ \;
    $\hspace*{2em}\FuncSty{colour}(G, \mathit{remaining} \cap \mathit{branchable}))$ \;
    \nl \While{$\FuncSty{length}(\mathit{colourClasses}) > 0$\label{outerloop}}{
        \nl \ForEach{$v \in \FuncSty{last}(\mathit{colourClasses}) \textnormal{~in reverse order}$\label{innerloop}}{
            \nl \lIf{$|\mathit{solution}| + \FuncSty{length}(\mathit{colourClasses}) \leq
            |\mathit{incumbent}|$}{\KwSty{return}\label{bound}}
            \nl \lIf{$v \notin \mathit{branchable} ~ \KwSty{and} ~ \mathit{solution} \ne
            \emptyset$}{\KwSty{return} \label{acceptbranchable}}
            \nl $\mathit{solution}' \gets \mathit{solution} \cup \{ v \} \label{addv} \label{addvtosolution} $\;
            \nl \lIf{$|\mathit{solution}'| > |\mathit{incumbent}|$}{$\mathit{incumbent} \gets
            \mathit{solution}'$ \label{newincumbent}}
            \nl $\mathit{branchable}' \gets \mathit{branchable} \cup \FuncSty{connected}(G, v)$ \label{addtobranchable} \;
            \nl $\mathit{remaining}' \gets \mathit{remaining} \cap \operatorname{N}(G, v)$ \label{filterremaining} \;
            \nl \lIf{$\mathit{remaining}' \neq \emptyset$}{$\FuncSty{search}(G, \mathit{solution}',
            \mathit{branchable}', \mathit{remaining}')$\label{recursesearch}}
         }
         \nl $\FuncSty{removeLast}(\mathit{colourClasses})$ \;
     }
}
\vspace{1ex}
\nl $\FuncSty{colour}$ :: (Graph $G$, Set $\mathit{uncoloured}$) $\rightarrow$ List of List of Vertex \;
\nl \Begin{
    \nl $\mathit{result} \gets []$ \;
    \nl \While{$\mathit{uncoloured} \ne \emptyset$\label{whileuncoloured}}{
        \nl $\mathit{colourClass} \gets []$ \label{newcolourclass} \;
        \nl $\mathit{colourable} \gets \mathit{uncoloured}$ \;
        \nl \While{$\mathit{colourable} \ne \emptyset$}{
            \nl $v \gets \textnormal{the first vertex from}~\mathit{uncoloured}~\textnormal{by
            static degree order from}~G$ \label{selectvtocolour} \;
            \nl $\FuncSty{append}(\mathit{colourClass}, v)$ \label{addtocolourclass} \;
            \nl $uncoloured \gets uncoloured \setminus \{ v \}$ \;
            \nl $\mathit{colourable} \gets \mathit{colourable} \setminus \operatorname{N}(G, v)$ \;
        }
        \nl $\FuncSty{append}(\mathit{result}, \mathit{colourClass})$ \;
    }
    \nl $\KwSty{return} ~ \mathit{result}$ \;
}
\vspace{1ex}
\nl $\FuncSty{connected}$ :: (Graph $G$, Vertex $v$) $\rightarrow$ Set \;
\nl \Begin{
    \nl $\KwSty{return} ~ \{ w \in G : \FuncSty{first}(w) \in \operatorname{N}(G, \FuncSty{first}(v)) \}$ \;
}
\vspace{1ex}
\nl $\FuncSty{associationGraph}$ :: (Graph $G_1$, Graph $G_2$) $\rightarrow$ Graph \;
\nl \Begin{
    \nl $V \gets \{ (v_1, v_2) \in G_1 \times G_2 : \FuncSty{compatible}(v_1, v_2) \}$ \;
    \nl $E \gets \{ ((v_1, v_2), (w_1, w_2)) \in V \times V : \FuncSty{compatible}((v_1, w_1), (v_2,
    w_2)) \}$ \;
    \nl $\KwSty{return} ~ (V, E)$ \;
}
\caption{An algorithm for MCCIS via an association graph.}
\label{algorithm:mccis}
\end{algorithm}

It is not possible to determine connectedness from a raw association graph. However, we can take a
maximum clique algorithm and mimic the branching strategy if we have access to the underlying graphs
and can determine the ``meaning'' of the association graph vertices.

Most modern maximum clique algorithms for dense graphs use some variation of greedy graph colouring
as a bound---the underlying observation is that each vertex in a clique must be given a different
colour in a colouring, so if we can colour a subset of vertices using $k$ colours then a maximum
clique in this subset has at most $k$ vertices. However, the colouring is also used as a branching
heuristic: vertices are selected in reverse from their colour classes in turn, starting with the
last colour class created. Because of this coupling of branching and the bound (which is important
in practice because it mimics a ``smallest domain first'' branching heuristic if colour classes are
viewed as variables \cite{DBLP:conf/cp/McCreeshP14}), if we were to select only a subset of vertices for
branching at each stage inside a clique algorithm, we would lose completeness. Thus we must adapt
the bound to take into account restricted branching.

In \cref{algorithm:mccis} we present a clique-inspired algorithm which finds an MCCIS via an
association graph. If the additional
branching restrictions are removed, the core of the algorithm is what Prosser
\cite{DBLP:journals/algorithms/Prosser12} calls the ``MCSa1'' variant of a series of algorithms due
to Tomita \textit{et al.}\
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10}, using a
bitset encoding approach due to San Segundo \textit{et al.}\
\cite{DBLP:journals/cor/SegundoRJ11,DBLP:journals/ol/SegundoMRH13} (and we refer the reader to these
papers for implementation details on how to use bitsets and other data structures to implement the
colouring stage with very low constant factors).

We begin by building the association graph (\lineref{buildassoc}). The main part of the algorithm
then works by building up candidate cliques in the $\mathit{solution}$ variable, by recursive calls
to the $\mathit{search}$ procedure---starting from the empty set (\lineref{firstsearch}), each
recursive subcall adds one vertex to $\mathit{solution}$ (\lineref{addv}) in such a way that
$\mathit{solution}$ is always a clique which corresponds to a connected common subgraph. The
$\mathit{remaining}$ set contains the set of vertices which are adjacent to every vertex in
$\mathit{solution}$, and which have not yet been accepted or rejected (and so initially it contains
every vertex). The main loops in the $\FuncSty{search}$ procedure
(\twolinesref{outerloop}{innerloop}) have the effect of iterating over each vertex in this set in a
particular order---each vertex $v$ is selected in turn, and then a recursive call is made to
consider the effects of including $v$ in $\mathit{solution}$ (\lineref{recursesearch}), followed by
the next iteration where $v$ is instead rejected. When $v$ is accepted, we add it to the new
$\mathit{solution'}$ (\lineref{addvtosolution}), and create a new $\mathit{remaining'}$ containing
only the vertices in $\mathit{remaining}$ which are adjacent to $v$ (\lineref{filterremaining}).

The $\mathit{branchable}$ set contains the set of association graph vertices which correspond to
vertices adjacent to an already-accepted vertex in the first input graph---in CP terms, it is the
set of assignments which could be made next which preserve connectedness. (Using only one of the two
input graphs is sufficient for correctness, and has the advantage that the $\FuncSty{connected}$
function may be implemented as a simple lookup into a precomputed array which maps each vertex in
the first input graph to a bitset.) At the top of search, this set is empty, and is not used (our
first vertex selection is special, and does not care about connectedness). At subsequent depths, we
may only accept vertices which are in this set, and if no such vertices remain then we return
immediately (\lineref{acceptbranchable}). When recursing, we extend $\mathit{branchable}$ with the
new vertices permitted by our acceptance of the branching $v$ (\lineref{addtobranchable}). Note that
we are assuming that inside the main loops, we encounter every vertex in $\mathit{remaining} \cap
\mathit{connected}$ before any vertex in $\mathit{remaining} \setminus \mathit{connected}$.

As we proceed, we keep track of the best solution we have found so far---this is stored in the
$\mathit{incumbent}$ variable (\twolinesref{incumbent}{newincumbent}). We use the incumbent,
together with a colour bound, to prune portions of the search space which cannot contain a better
solution. The colour bound operates as follows: at each entry to the $\FuncSty{search}$ procedure,
we produce a greedy colouring of the vertices in $\mathit{remaining}$ (\lineref{makecolours}). This
greedy colouring gives us a list of colour classes, each of which is a list of pairwise non-adjacent
vertices. The two loops (\twolinesref{outerloop}{innerloop}) then iterate over each colour class,
from last to first, and then over each vertex in that colour class, again from last to first.
(Rather than actually using a list of lists and removing items, this process should be implemented
using a pair of immutable flat arrays. This technique is described elsewhere
\cite{DBLP:conf/cp/McCreeshP14}, so we do not discuss it here.) Finally, if at any point the number
of remaining colour classes plus the number of vertices currently present in $\mathit{solution}$ is
not strictly greater than the size of the incumbent, then we may backtrack immediately
(\lineref{bound}).

Finally, we describe the colouring process. In conventional clique algorithms, a simple greedy
sequential colouring is used (possibly with the help of previous colourings to reduce the
computational cost \cite{DBLP:conf/lion/NikolaevBS15}, and possibly with shortcuts taken for certain
vertices \cite{DBLP:journals/cor/SegundoT14}, and possibly followed by a repair step to
improve the colouring \cite{DBLP:conf/walcom/TomitaSHTW10}, or stronger bounding rules based upon
MaxSAT inference \cite{DBLP:conf/ictai/LiFX13,DBLP:conf/lion/LiJX15,DBLP:journals/cor/SegundoNB15}).
Such colourings will not give us the required property that vertices in $\mathit{remaining} \cap
\mathit{connected}$ come last (so they are selected first by the reverse branching order). Thus we
produce two greedy sequential colourings, first considering the non-branching vertices in
$\mathit{remaining} \setminus \mathit{connected}$, followed by the branching vertices, and
concatenate them (\lineref{makecolours}). This produces a valid colouring, since we do not merge any
colour classes between the two stages, although it may use more colours than a single colouring
would.

Our $\FuncSty{colour}$ procedure is the same as the bit-parallel algorithm introduced by San Segundo
\textit{et al.}\ \cite{DBLP:journals/cor/SegundoRJ11}. While we have remaining vertices to colour
(\lineref{whileuncoloured}), we start a new colour class (\lineref{newcolourclass}), and then
repeatedly pick a legal vertex to add to that colour class (\lineref{addtocolourclass}).  The
selection of the next vertex to colour (\lineref{selectvtocolour}) may be performed efficiently in
hardware if the association graph is permuted to be in degree order at the top of search. (Other
initial vertex orderings have been considered on general clique problems
\cite{DBLP:journals/algorithms/Prosser12,DBLP:conf/lion/SegundoLB14}; it is possible that special
properties of the association graph could be exploited in this step.)
